---
title: "Sample Information and Normalisation Setup"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(DiffBind)
library(magrittr)
library(DESeq2)
library(scales)
```

# Overview
# Constructs sample metadata, loads BAM read counts, computes spike-in and
# greenlist normalisation factors, and generates random group comparisons
# for downstream DiffBind analysis.

---

# 1. Define directories

```{r directories}
working_dir     <- "path/to/analysis"                          # adjust to your system
results_dir     <- file.path(working_dir, "results")
bam_hg38_dir    <- file.path(results_dir, "bam_files/final_bams/hg38")
peaks_dir       <- file.path(results_dir, "macs2/macs2_q0.01/with_no_neg_control/narrowPeak")
greenlist_dir   <- file.path(working_dir, "R_analysis_final/01.Greenlist_norm_calcs")
out_dir         <- file.path(working_dir, "R_analysis_final/02.Sample_information/data")
```

---

# 2. Build sample information table

```{r sample-info}
# Collect BAM and peak file paths
hg38_bam_files <- file.path(
    bam_hg38_dir,
    list.files(bam_hg38_dir, pattern = "_OTHER_hg38_nodm6_chrmrm_blrm_duprm_unmappedrm_multimaprm\\.bam$")
)

peak_files <- file.path(
    peaks_dir,
    list.files(peaks_dir, pattern = "\\.narrowPeak$")
)

# Helper: strip path prefixes and filename suffixes to get clean sample names
clean_names <- function(paths, prefix, suffix) {
    paths |> gsub(prefix, "", x = _) |> gsub(suffix, "", x = _)
}

bam_names  <- clean_names(hg38_bam_files, paste0(bam_hg38_dir, "/"),
                          "_Homo_sapiens_OTHER_hg38_nodm6_chrmrm_blrm_duprm_unmappedrm_multimaprm.bam")
peak_names <- clean_names(peak_files, paste0(peaks_dir, "/"), "_peaks.narrowPeak")

# Verify BAM and peak files are in matching order
stopifnot("BAM and peak files are not in matching order" = all(peak_names == bam_names))

# Clean up sample IDs
clean_sample_ids <- function(x) {
    x |>
        gsub("0_6",              "0.6",   x = _) |>
        gsub("2_5",              "2.5",   x = _) |>
        gsub("7_5",              "7.5",   x = _) |>
        gsub("_Homo_sapiens_OTHER_hg38_nodm6", "", x = _) |>
        gsub("_PS_HsDm_CTCF",   "",      x = _) |>
        gsub("_Cut_and_Run",     "",      x = _) |>
        gsub("0602_",            "",      x = _) |>
        gsub("_GSM",             "-GSM",  x = _)
}

SampleID <- clean_sample_ids(bam_names)

# Assemble sample table
sample_information <- data.frame(
    SampleID   = SampleID,
    bamReads   = hg38_bam_files,
    Peaks      = peak_files,
    PeakCaller = "narrow",     # narrow peak caller appropriate for transcription factors
    CellLine   = "K562"
) |>
    separate(SampleID, into = c("Antibody", "Condition"), sep = "_", remove = FALSE)
```

---

# 3. Load and join BAM count summary (spike-in normalisation)

```{r bam-counts}
summary_bam_counts <- read_tsv(file.path(results_dir, "summary_hg38_dm6_bam_counts.txt"))

# Column names are stored space-separated in the first column header — reshape
col_names <- names(summary_bam_counts) |> strsplit(split = " ") |> unlist()
summary_bam_counts <- separate(summary_bam_counts, 1, into = col_names, sep = " ")

# Clean sample IDs to match sample_information
summary_bam_counts$basename <- clean_sample_ids(summary_bam_counts$basename)

summary_bam_counts <- summary_bam_counts |>
    separate(basename, into = c("Antibody", "Condition"), sep = "_", remove = FALSE)

# Convert BAM count columns to numeric
bam_count_cols <- grepl("bam_count", colnames(summary_bam_counts))
summary_bam_counts[, bam_count_cols] <- lapply(
    summary_bam_counts[, bam_count_cols], as.numeric
)

# Spike-in (dm6) normalisation factor: scale relative to mean dm6 read depth
summary_bam_counts$dm6_reads_norm_scale <-
    summary_bam_counts$dm6_intermediate_bam_counts /
    mean(summary_bam_counts$dm6_intermediate_bam_counts)

# Join to sample table and compute library size normaliser
sample_information <- full_join(sample_information, summary_bam_counts, by = c("Antibody", "Condition"))
sample_information$lib_size_normaliser_calc <-
    sample_information$hg38_final_bam_counts /
    mean(sample_information$hg38_final_bam_counts)
```

---

# 4. Filter samples

```{r filter-samples}
# Remove CTCF-disrupted control (not part of time-course comparisons)
# Remove 5s time point (low alignment rate)
sample_information <- sample_information |>
    filter(!grepl("disrupted", SampleID),
           !grepl("5s",        SampleID))
```

---

# 5. Generate random group comparisons

```{r random-comparisons}
set.seed(3)
nruns <- 30

rand_combos <- data.frame(matrix(nrow = nruns, ncol = 6))
names(rand_combos) <- c("A1", "A2", "A3", "B1", "B2", "B3")

sample_information_randn <- vector("list", nruns)

for (i in seq_len(nruns)) {
    idx <- sample(nrow(sample_information), 6)
    rand_combos[i, ] <- t(idx)
    sample_information_randn[[i]] <- sample_information[idx, ]
    sample_information_randn[[i]]$Factor    <- c("A", "A", "A", "B", "B", "B")
    sample_information_randn[[i]]$Replicate <- c(1, 2, 3, 1, 2, 3)
}
```

---

# 6. Compute greenlist normalisation factors per comparison

```{r greenlist-norm}
# Load greenlist region counts (rows = genomic regions, cols = samples)
counts_raw <- read.delim(
    file.path(greenlist_dir, "glist_quant.tsv"),
    header = TRUE, row.names = 1, check.names = FALSE
)

# Prepend full BAM path so columns match sample_information$bamReads
colnames(counts_raw) <- file.path(bam_hg38_dir, paste0(colnames(counts_raw), ".bam"))

for (i in seq_len(nruns)) {
    # Subset to samples in this comparison (preserving order)
    counts_sub <- counts_raw[, match(sample_information_randn[[i]]$bamReads,
                                     colnames(counts_raw))]

    # Estimate size factors using DESeq2's median-of-ratios method
    sf <- estimateSizeFactorsForMatrix(counts_sub) |>
        as.data.frame() |>
        setNames("sf")
    sf$greenlist_normalizer <- 1 / sf$sf
    sf$bamReads <- rownames(sf)

    sample_information_randn[[i]] <- full_join(sample_information_randn[[i]], sf,
                                               by = "bamReads")
}
```

---

# 7. Save outputs

```{r save-outputs}
# Summary table of random comparisons
comparison_table <- tibble(
    run        = seq_len(nruns),
    comparison = sapply(seq_len(nruns), function(i) {
        b <- paste(sample_information_randn[[i]]$Condition[sample_information_randn[[i]]$Factor == "B"],
                   collapse = ", ")
        a <- paste(sample_information_randn[[i]]$Condition[sample_information_randn[[i]]$Factor == "A"],
                   collapse = ", ")
        paste(b, "vs", a) |> gsub("million", "mill", x = _)
    })
)

write.csv(comparison_table,
          file = file.path(out_dir, "list_of_comparisons.csv"),
          row.names = FALSE)

saveRDS(sample_information,
        file = file.path(out_dir, "sample_information_all.rds"))

saveRDS(sample_information_randn,
        file = file.path(out_dir, "sample_information_randn.rds"))

save.image(file = file.path(out_dir, "sample_information_workspace.RData"))
```
